Choosing the Choice: Reflections on Modelling Decisions and Behaviour in Demographic Agent-Based Models

### 2016-06-27 article.tex
Minor tweaks

As demonstrated throughout this Special Issue, agent-based models have become important tools with which demographers can study mechanisms of social reality, enhancing the analysis of population processes with multi-level influences, feedback effects, the presence of networks and interactions. Still, the related methodology remains under development: in particular, as evident in the nascent demographic agent-based literature, there is hardly any discussion on operationalising the actual decision making processes, through which the agents could acquire some more agency beyond simply reacting to simulated stimuli.

@@

This section stresses the axiomatic status of models of choice in agent-based simulation, and discusses the numerous interfaces between choice and context at which modelling decisions must be made.

@@

Neither approach is a complete account. While mimicry seems to be fundamental to many aspects of human behaviour \citep{Chartrand1999}, we are not simply impersonators and instead interpret the actions of others. Absent telepathy however, we can only draw inferences based on observation of their actions\footnote{Which may well include communicating about their internal state.}. A potential approach to this might be one similar to that of \citet{Jern2011a} who propose an inverted generative model to explain how people might translate observed behaviour into preferences, although this presages a utility type mechanism for decision making. This also has implications for more high level models of the interaction between context and decision making process, for example that of \citet{Ben-Akiva2012}, where there are explicit models of the interaction between context and choice. An inverted model which does not sufficiently capture the context in which an agent's decision making took place, i.e. the observing agent is underinformed about inputs to the other's decision process, or constraints on their actions, may well lead to faulty inferences about what drove their behaviour. Although this is realistic, in that our own inferences about the motivations of others are not always accurate, it may not be desirable in all models.

@@

\citeauthor{McKenzie2013} also suggest that overweighting of negative experiences plays a role, which demonstrates the importance of considering how information is incorporated into the decision process. In this case the suggestion is that another cognitive bias, loss aversion, is a salient feature. This could be reflected by using a decision model explicitly incorporates loss aversion, for example Prospect Theory \citep{Kahneman1979}, or Cumulative Prospect Theory \citep{Tversky1992}. Both variants of Prospect Theory are intended to mimic human decision making, by distorting the perception of high and low probabilities towards certainty, and shifting the value of gains and losses such that large gains are progressively less significant, and losses more so. 
The wider implication is that social information flow may be distorted at the source, or through interpretation, and that agent-based models are uniquely suited to capture this.  Other approaches to decision models are available, for example heuristic methods \citep{Gigerenzer1996}, which emphasise the bounded rationality of human decision making, and Bayesian approaches, which address learning as a component of the decision process. In an agent-based modelling context, \citet{Gray2016} have contrasted Bayesian, heuristic, and Cumulative Prospect Theory approaches in a simulation of alcohol misuse disclosure among pregnant women.

@@

The diversity of approaches to this single issue is reflective of the open ended nature of agent-based modelling. A recurring feature however, is that an appreciation is needed of the multiple roles of contexts in decision making. Interpersonal context is a critical feature of information transmission, but also plays a role in the subjective value of outcomes. This is a challenge, since it is not clear that the two are readily separable, and our understanding of the context is incomplete. 
However the opportunity this presents is to use simulation to expand our understanding both of the role of social processes, and their mechanics.

@@

There are many good reasons why ignoring time in models of decision making might be valid or even desirable (and, as modellers, we have often done so). As was noted above, the context and research question must of course inform modelling decisions. It may be that the decisions in question have only short term consequences, and so a detailed consideration of how agents treat time in their decision making be completely superfluous.

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behavior to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem at hand. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.

>>>>

As demonstrated throughout this Special Issue, agent-based models have become important tools with which demographers can study mechanisms of social reality, enhancing the analysis of population processes with multi-level influences, feedback effects, the presence of networks and interactions. Still, the related methodology remains under development: in particular, as evident in the nascent demographic agent-based literature, there is hardly any discussion on operationalising the actual decision making processes, through which the agents could acquire some more agency beyond reflexive reactions to simulated stimuli.

@@

This section stresses the axiomatic status of the current models of choice in agent-based simulation, and discusses the numerous interfaces between choice and context at which modelling decisions must be made.

@@

Neither approach is a complete account. While mimicry seems to be fundamental to many aspects of human behaviour \citep{Chartrand1999}, we are not simply impersonators and instead interpret the actions of others. Absent telepathy however, we can only draw inferences based on observation of their actions\footnote{Which may well include communicating about their internal state.}. A potential approach to this might be one similar to that of \citet{Jern2011a} who propose an inverted generative model\footnote{A generative model produces decisions, given preferences over outcomes. The inversion provides estimates of preferences, given observed decisions.} to explain how people might translate observed behaviour into preferences, although this presages a utility type mechanism for decision making. This also has implications for more high level models of the interaction between context and decision making process, for example that of \citet{Ben-Akiva2012}, where there are explicit models of the interaction between context and choice. An inverted model which does not sufficiently capture the context in which an agent's decision making took place, i.e. the observing agent is underinformed about inputs to the other's decision process, or constraints on their actions, may well lead to faulty inferences about what drove their behaviour. Although this is realistic, in that our own inferences about the motivations of others are not always accurate, it may not be desirable in all models.

@@

\citeauthor{McKenzie2013} also suggest that overweighting of negative experiences plays a role, which demonstrates the importance of considering how information is incorporated into the decision process. In this case the suggestion is that another cognitive bias, loss aversion, is a salient feature. This could be reflected by using a decision model explicitly incorporates loss aversion, for example Prospect Theory \citep{Kahneman1979}, or Cumulative Prospect Theory \citep{Tversky1992}. Both variants of Prospect Theory are intended to mimic human decision making, by distorting the perception of high and low probabilities towards certainty, and shifting the value of gains and losses such that large gains are progressively less significant, and losses more so. 
The wider implication is that social information flow may be distorted at the source, or through interpretation, and that agent-based models are uniquely suited to capture this.  Other approaches to decision models are available, for example heuristic methods \citep{Gigerenzer1996}, which emphasise the bounded rationality of human decision making, and Bayesian approaches, which address learning as a component of the decision process \citep{Robbins1964}. In an agent-based modelling context, \citet{Gray2016} have contrasted Bayesian, heuristic, and Cumulative Prospect Theory approaches in a simulation of alcohol misuse disclosure among pregnant women.

@@

The diversity of approaches to this single issue is reflective of the open ended nature of agent-based modelling. A recurring feature however, is that an appreciation is needed of the multiple roles of contexts in decision making. Interpersonal context, for example, is a critical feature of information transmission, but also plays a role in the subjective value of outcomes. This is a challenge, since it is not clear that the two are readily separable, and our understanding of interpersonal context is incomplete.
However this presents us with great opportunity, in using simulation to expand our understanding both of the role of social processes, and their mechanics; and in following the example of \citeauthor{Tversky1992} by forging inter-disciplinary collaborations with experts in the micro-domain of behaviour.

@@

There are many good reasons why ignoring time in models of decision making might be valid or even desirable (and, as modellers, we have often done so). As was noted above, the context and research question must of course inform modelling decisions. It may be that the decisions in question have only short-term consequences, and so a detailed consideration of how agents treat time in their decision making be completely superfluous.

@@

So, in the light of the above arguments, how to choose the appropriate representation of the human decisions and behavior to be used in the agent-based modelling of demographic processes? First, we need to remember that agent-based models are a very convenient tool, with which we can integrate behavioural theory with social theory and data. This implies that – as suggested in the Manifesto of computational social science \citep{Conte} – such models would ideally need to collect additional information on human decision making through bespoke cognitive experiments, instead of merely relying on hypotheses or assumptions \citep{Courgeau}. In this way, the choice of a particular choice model will become an empirical issue, but specific to a given context and the decision problem of interest. Following this approach would allow for including empirical micro-foundations in computational demographic models in an open and transparent way – and provide a natural, bottom-up and empirical way if evaluating the suitability of particular choice theories for the problem at hand.
